---
title: "rehoming"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)
```

## Introduction to Data

Information about dogs and horses available for adoption was scraped from the Blue Cross and and Dog's Trust websites into a SQL Server database on Microsoft Azure. Scraping scripts were written in Python's Scrapy framework. The spiders were within a Docker image which was run automatically every 24 hours by an Azure Logic App. Data were stored in an Azure SQL Server database. This database had three tables: `centres`, `adoptions`, and `pets`. Each of the tables was obtained in delimited format by querying the database through the bcp command line utility. 

The `centres` table contains one row per encountered centre. 30 different rehoming centres were encountered. 

```{r}
centres <- read_csv(here("data", "centres.csv"), 
                    col_types = "c",
                    col_names = c("centre_name"))
head(centres)
nrow(centres)
```

Each pet present on the site on a given day gets a new row in the `adoptions` table. Pets may continue to be advertised even when reserved, and when this is the case their value in the `reserved` column will be `TRUE`. There are 151,545 records in the `adoptions` table, spanning 386 days. 

```{r}
adoptions <- read_csv(here("data", "adoptions.csv"),
                      col_names = c("pet_id", "reserved", "date_scraped"),
                      col_types = "clD") %>%
  distinct()
head(adoptions)
nrow(adoptions)
length(unique(adoptions$date_scraped))
```

Each row in the `pets` table contains details on a single animal. Data was scraped on 6,176 unique pets. There were three malformed rows in this data, most likely due to a new line 
character being contained within a pet description. 

bcp "SELECT * FROM pets" queryout "C:\Users\s1533194\pets.csv" -c -S anailis-personal.database.windows.net -U EF1079 -d rehoming -t "|" -r '#'
$f = (gc .\pets.csv) -join ''
$f.split("#") > tmp.csv
(gc .\tmp.csv) -replace '`0','' | Out-File -Encoding UTF8 final.csv

```{r}
pets <- read_delim(here("data", "pets.csv"), delim = "|",
                   col_names = c("pet_id", "species", "pet_type", "sex",
                                 "age", "centre_name", "info", "description",
                                 "height"),
                   col_types = c("c")) %>%
  filter(!is.na(species))
nrow(pets)
length(unique(pets$pet_id))
head(pets)
```

A sanity check: every pet should be represented at least once in the `adoptions` dataframe. 

```{r}
sum(pets$pet_id %in% adoptions$pet_id) == length(unique(pets$pet_id))
```

## Initial Data Exploration 

### Data Sources

Dogs Trust (DT) IDs begin with the prefix `DT`, whereas Blue Cross (BC) IDs are just numbers. 
We can use this to create columns that indicate data source. 

```{r}
pets$data_source <- ifelse(str_detect(pets$pet_id, "DT"), "DT", "BC")
adoptions$data_source <- ifelse(str_detect(adoptions$pet_id, "DT"), "DT", "BC")
```


Blue Cross was scraped between January and early March 2021 (63 days). After this, 
the data source was changed to Dogs Trust, which was scraped between early March 2021 and 
Februrary 2022 (337 days). 

For every day of Blue Cross scraping, scraping was successful (63/63). However, there were 13 
days where Dogs Trust scraping failed (324/337).

```{r}
adoptions %>%
  group_by(data_source) %>%
  summarise(
    first_scrape = min(date_scraped),
    last_script = max(date_scraped),
    scraping_length = max(date_scraped) - min(date_scraped),
    total_dates_scraped = length(unique(date_scraped))
  )
```

As a consequence of the longer scrape length, more pets come from Dogs Trust than Blue Cross. 

```{r}
pets %>%
  group_by(data_source) %>%
  count()
```

Blue Cross lists horses and dogs, whereas Dogs Trust only lists dogs. 

```{r}
pets %>%
  group_by(species, data_source) %>%
  count()
```

### Listed Dates

Limiting to just the Dogs Trust data, we can look at the numbers of dogs listed on each 
day that there was a successful scrape. 

```{r}
adoptions %>%
  filter(data_source == "DT") %>%
  group_by(date_scraped) %>%
  count() %>%
ggplot(aes(x = date_scraped, y = n)) + 
  geom_point() +
  labs(
    y = "Dogs Listed",
    x = "Date"
  ) +
  theme_bw()
```

Let's select a random pet from the `pets` dataframe and follow the journey its data would have
taken. Let's examine DT1021335. This pet would have been entered into `pets` the first time 
this ID was encountered. It would have also been entered into the `adoptions` table for 
the first time. We can see that the first time this pet was encountered was on 11/08/2021.  
Every subsequent day pet DT1021335 was encountered, it would get yet another entry in the 
`adoptions` table. The last entry for DT1021335 was on 11/10/2021, 61 days after it was 
first observed. After this, pet DT1021335 was not observed again. This means that 
DT1021335 was advertised as available for adoption for a total of 61 days. 

```{r}
pets %>%
  filter(pet_id == "DT1021335")

adoptions %>%
  filter(pet_id == "DT1021335") %>%
  arrange(date_scraped)
```

The first and last date observed for each pet is obtained and added to the `pets` table. 

```{r}
first_observed <- adoptions %>% 
  select(-data_source) %>%
  group_by(pet_id) %>%
  slice_min(date_scraped)
nrow(first_observed) == nrow(pets)

last_observed <- adoptions %>%
  select(-data_source) %>%
  group_by(pet_id) %>%
  slice_max(date_scraped)
nrow(last_observed) == nrow(pets)

pets <- pets %>%
  left_join(last_observed, by = "pet_id") %>%
  left_join(first_observed, by = "pet_id",
            suffix = c(".last", ".first"))
nrow(pets)
```

From this, the number of days a pet was listed for adoption can be calculated. 

```{r}
pets$days_listed <- pets$date_scraped.last - pets$date_scraped.first + 1
```

Most dogs at Dogs Trust are listed only for a few days, but an unfortunate few are listed 
for hundreds of days. 

```{r}
pets %>%
  filter(data_source == "DT") %>%
ggplot(aes(x = days_listed)) + 
  geom_histogram(colour = "black") +
  labs(
    x = "Days Listed",
    y = "Number of Dogs"
  ) +
  theme_bw()
```
However, on the Blue Cross website pets may remain listed even when they are reserved. 
This is why the  `adoptions` table has a `reserved` column to indicate whether the pet is
listed as reserved. 
So we need to add one more piece of information: the first date at which a pet was listed 
as reserved (if at all). 

```{r}
first_reserved <- adoptions %>%
  filter(reserved) %>%
  group_by(pet_id) %>%
  slice_min(date_scraped) %>%
  select(pet_id, date_scraped) %>%
  rename(date_scraped.reserved = date_scraped)

pets <- pets %>%
  left_join(first_reserved, by = "pet_id")
```

The days until a Blue Cross pet is reserved are calculated. 

```{r}
pets$days_until_reserved <- pets$date_scraped.reserved - pets$date_scraped.first
```

### Species 

Most of the animals represented are dogs. 

```{r}
pets %>%
  group_by(species) %>%
  count()
```

Dogs never have an entry in their height columns, whereas only one of the 63 horses 
identified had a missing height field. 

```{r}
pets %>% 
  group_by(species) %>%
  summarise(
    missing_height = sum(is.na(height)),
    prop_missing_height = sum(is.na(height))/n()
    )
```

### Age

Age is represented differently for Blue Cross and Dogs Trust. Blue Cross has an exact age (e.g. 5 years 2 months), whereas Dogs Trust uses a system of 6 age categories. 

```{r}
pets %>%
  filter(data_source == "BC") %>%
  select(age) %>%
  distinct() %>%
  head()

pets %>%
  filter(data_source == "DT") %>%
  select(age) %>%
  distinct()
```

I separate the Dogs Trust data into its own dataframe, `dogstrust_pets`, so I can turn 
the age categories into factor levels. 

```{r}
dt_age_levels <- c("0 to 6 Months", "6 to 12 Months", "1 to 2 Years",
            "2 to 5 Years", "5 to 7 Years", "8 and over")

dogstrust_pets <- pets %>%
  filter(data_source == "DT") %>%
  mutate(age = fct_relevel(age, dt_age_levels))
```

Most pets are in the 2 to 5 year age bracket. Notably there are a large number of senior dogs,
over 8 years old. 

```{r}
dogstrust_pets %>%
ggplot(aes(x = age)) + 
  geom_bar(colour = "black") + 
  labs(
    x = "Age Group",
    y = "Number of Dogs"
  ) + 
  theme_bw()
```

In all age groups, there were a small number of dogs listed for a very long time. However, 
the violin plot does demonstrate that puppies (dogs between the ages of 0 and 6 months), 
were less likely to have these extreme listing times compared to dogs in other age groups. 

```{r}
dogstrust_pets %>%
ggplot(aes(x = age, y = days_listed)) + 
  geom_violin(fill = "grey") + 
  labs(
    x = "Age Group",
    y = "Days Listed"
  ) + 
  theme_bw()
```

The median time a puppy between 0 and 6 months was listed was just one day. 

```{r}
dogstrust_pets %>%
  group_by(age) %>%
  summarise(
    mean_days_listed = mean(days_listed),
    median_days_listed = median(days_listed),
    min_days_listed = min(days_listed),
    max_days_listed = max(days_listed)
  )

dogstrust_pets %>%
  filter(age == "0 to 6 Months" & days_listed == 338) %>%
  select(description)
```

### Sex

Male dogs made up 60.6% of all dogs listed, and 60.5% of all dogs listed on Dogs Trust. 

```{r}
pets %>%
  group_by(species, sex) %>%
  count()

dogstrust_pets %>%
  group_by(sex) %>%
  count()
```

### Breed

The most commonly listed breed is simply "a crossbreed". The next most common breeds are JRTs, lurchers, and greyhounds. 

```{r}
dogstrust_pets %>%
  group_by(pet_type) %>%
  count() %>%
  arrange(desc(n))
```


### Special Requirements 

Pets with the string Medic/medic in their description may have medical needs. 

```{r}
dogstrust_pets$medical_needs <- str_detect(dogstrust_pets$description, "[m|M]edic")
```

The number of dogs with Medic/medic in their description is greater for age groups 
representing older dogs, and is highest for the senior dog category (8 and over).

```{r}
dogstrust_pets %>%
  filter(medical_needs) %>%
ggplot(aes(x = age)) + 
  geom_bar(colour = "black") + 
  labs(
    y = "Number of Dogs",
    x = "Age Group"
  ) +
  theme_bw()
```

```{r}
dogstrust_pets %>%
  filter(age == "0 to 6 Months" & medical_needs)

dogstrust_pets %>%
  filter(age == "0 to 6 Months") %>%
  arrange(desc(days_listed))
```

### Sentiment 

Look for commonly occuring words in descriptions:
e.g. "Medical", "medication" 
